\input{../../common.tex}

\addbibresource{root.bib}

\title{Practical TLA\textsuperscript{+}}

\author{Jan Macháček}

\begin{document}

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
        The only way for me to find out what my lively, but chaotic in-laws have planned for the weekend is to phone the grandmother's house to find out who is there; and then call or text the people on the list to find the actual answers. (Don't for one minute think that it's the grandmother's job to get people to the phone!) The nature of the calls and texts is that some reply immediately with their answers; some reply, but haven't made a decision yet; some don't respond to calls, but will eventually text back; some have their phones off or out of batteries. The challenge is to contact everyone on the list within 10 minutes; or else go back to the start of the process!

        An actor system using the Akka toolkit seems like a perfect fit for the implementation: one actor as an aggregate of what all in-laws are doing coordinating an actor-per-in-law to handle the person's plans, and an actor for each communication method: the mobile phone, messenger app, etc. Perfect! The happy-day scenario, where everyone responds in time, will work and will be efficient; but what about the case where there are time-outs, failed calls, undelivered text messages? 

        This paper shows how to use TLA\textsuperscript{+} to construct a model of the various states the system will be in, verify the invariants that should hold in each state; but also discover traces where this model fails. Once the model is verified, it can be turned to Akka and Scala code.
    \end{abstract}
    \bigskip
  \end{@twocolumnfalse}
]

\section{Lively, but chaotic family system}

\printbibliography

\end{document}
